---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
```{r packages, include = FALSE}
library(pensionviewr)
library(tidyverse)
library(kableExtra)
library(reasontheme)
```
# pensionviewr

The goal of `pensionviewr` is to simplify the process of gathering and visualizing public pension plan data from the Reason pension database. This repo contains the functions of the `pensionviewr` package, which once installed locally, provides helpful functions for creating and exporting graphics made in ggplot in the style used by the Reason Pension Integrity Project team.

## Installing pensionviewr

`pensionviewr` is not on CRAN, so you will have to install it directly from Github using `devtools`. 

`devtools` calls the GitHub API (Application Programming Interface). And this requires that you authenticate yourself in a different way than the usual username and password. For the GitHub API, we use a personal access token (PAT), which is a convention followed by many other APIs.

### Step-by-step

source: https://happygitwithr.com/github-pat.html

Login to your GitHub account. Go to https://github.com/settings/tokens.

“Generate new token”. Give it a nickname that reminds you of the intended purpose, e.g., “devtools”.

Pick scopes that confer the privileges you need. When in doubt, check the repo scope, which is required for typical devtools operations. There used to be a set of default scopes, but that no longer appears to be true.

After you click “Generate token”, the token will be displayed. It’s a string of 40 random letters and digits. This is the last time you will see it SO COPY IT TO THE CLIPBOARD. Leave this window open until you’re done. If you somehow goof this up, just generate a new one and try again.

Put your PAT in your .Renviron file. Have a line that looks like this:

```
GITHUB_PAT=8c70fd8419398999c9ac5bacf3192882193cadf2
```

but with your PAT instead of mine. Don’t worry, I’ve revoked this one!

`.Renviron` is a hidden file that lives in your home directory. Not sure where that is? The easiest way to find and edit `.Renviron` is with a function from the `usethis` package. In R, do:

```
usethis::edit_r_environ()
```

Your `.Renviron` file should pop up in your editor. Add your GITHUB_PAT as above, save and close it.

Put a line break at the end.

Restart R (Session > Restart R in the RStudio menu bar), as environment variables are loaded from `.Renviron` only at the start of an R session. Check that the PAT is now available like so:

```
Sys.getenv("GITHUB_PAT")
```

You should see your PAT print to screen.

Now commands you run from the `devtools` package, which consults `GITHUB_PAT` by default, will be able to access private GitHub repositories to which you have access.

If you do not have the `devtools` package installed, you will have to run the first line in the code below as well. 

``` r
install.packages('devtools')
devtools::install_github("ReasonFoundation/pensionviewr")
```

## Using the functions:

The package has four functions for data pulling and preparation: `planList()`, `pullData()`, `loadData()`, and `selectedData()`.

The package has four functions for plots: `glPlot()`, `linePlot()`, `debtPlot()`, and `savePlot()`.

A basic explanation and summary here:

### `planList()`

1. `planList()`: returns a stripped down list of the pension plans in the database along with their state and the internal databse id.

Example of how it is used in a standard workflow:

```{r planList}
pl <- planList()
pl %>% 
  head() %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")
```

### `pullData()`

2. `pullData()`:  pulls the data for a specified plan from the Reason pension databse. `pullData` has two arguments:
`pullData(pl, plan_name)`

* `pl`: A datafram containing the list of plan names, states, and ids in the form produced by the `planList()` function.
* `plan_name`: A string enclosed in quotation marks containing a plan name as it is listed in the Reason pension database.

Example of how it is used in a standard workflow:

The next step would be to load the data for the specific plan of interest. Let's use the Vermont State Retirement System as an example. Let's first see what plans in Vermont are available:

```{r Kansas}
VT <- pl %>% filter(state == 'Vermont')
VT %>% 
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")
```

The full plan name we are interested in is there listed as "Vermon State Retirement System". We can pull the data for it now:

```{r pullData}
vtsrs_data <- pullData(pl, plan_name = "Vermont State Retirement System")
vtsrs_data %>% 
  head() %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")
```

### `loadData`

3. `loadData`: loads the data for a specified plan from an Excel file. `loadData` has one argument:

`loadData(file_name)`

* `file_name`: A string enclosed in quotation marks containing a file name with path of a pension plan Excel data file.

```
data_from_file <- loadData('data/NorthCarolina_PensionDatabase_TSERS.xlsx')
```

### `selectedData()`

4. `selectedData()`: selects the only the variables used in historical analyses. `selectedData` has one argument, `wide_data`, that is required:

`selectedData(wide_data)`

* `wide_data`: a datasource in wide format

Back to the Kansas Public Employees' example. That is a lot of variables. The `selectedData()` function selects only a handful of needed variables:

```{r selectedData}
df <- selectedData(vtsrs_data)
df %>% 
  head() %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")
```

### `glPlot()`

5. `glPlot()`: creates the 'Gain/Loss' plot using a CSV file as an input. glPlot has two arguments:

`glPlot(filename, ylab_unit)`

* `filename`: a csv (comma separated value) file containing columns of gain loss category names with one row of values.
* `ylab_unit`: a string contained within quotation marks containing th y-axis label unit. The default value is "Billions"

Example of how it is used in a standard workflow:

```
filename <- "data/GainLoss_data.csv"
glPlot(filename)
```

### `linePlot()`

6. `linePlot()`: creates a plot comparing two variables, such as ADEC vs. Actual contributions. `linePlot()` has six arguments, with `data` being required:

`linePlot(data, .var1, .var2, labelY, label1, label2)`

* `data` a dataframe produced by the selectedData function or in the same format.
* `.var1` The name of the first variable to plat, default is adec_contribution_rates.
* `.var2` The name of the second variable to plot, default if actual_contribution_rates.
* `labelY` A label for the Y-axis.
* `label1` A label for the first variable.
* `label2` A label for the second variable.

```{r contributions}
linePlot(df)
```

### `debtPlot()`

8. `debtPlot()`: creates the "History of Volatile Solvency" or "Mountain of Debt" chart. `debtPlot` takes one argument:

`debtPlot(data)`

* `data`: a dataframe produced by the `selectedData()` function or in the same format containing year, uaal, funded ratio columns.

```{r plot1, echo = FALSE}
debt_plot <- debtPlot(df)
debt_plot
```

### `savePlot()`

9. `savePlot()`: adds a source and save ggplot chart. `savePlot` takes five arguments:
`savePlot(plot = myplot, source = "The source for my data", save_filepath = "filename_that_my_plot_should_be_saved_to.png", width_pixels = 648, height_pixels = 384.48)`

* `plot`: The variable name of the plot you have created that you want to format and save
* `source`: The text you want to come after the text 'Source:' in the bottom left hand side of your side
* `save_filepath`: Exact filepath that you want the plot to be saved to
* `width_pixels`: Width in pixels that you want to save your chart to - defaults to 648
* `height_pixels`: Height in pixels that you want to save your chart to - defaults to 384.48

```
savePlot(debt_plot, source = "Source: KPERS", save_filepath = "output/test.png")
```

The BBC has created a wonderful data journalism cookbook for R graphics located here:
https://bbc.github.io/rcookbook/




